Prevailing scenarios of functional change in Anthropocene bird and mammal communities

### 2023-06-23 paper/GEB Submission/manuscript_GEB.Rmd

  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics. Using generalized linear mixed models we estimated general trends in those metrics and trends for individual studies.

  Results: We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover. The lack of trend held even after correcting for changes in species richness. At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.

@@

The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
Still, there is general consensus on a few local-scale patterns that seem to be characteristic of the anthropocene, including the prevalence of species turnover across communities and patterns of both richness increase and decrease in different taxa and biomes [@dornelas2023].
Understanding the implications of change in species assemblage for ecosystem stability and function is a critical next step in moving forward the biodiversity change conversation.

@@

Here we perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
We distinguish between three possible scenarios of functional change: 1) significant functional change accompanied by changes in species-based metrics, 2) maintenance of functional structure despite significant species changes, 3) maintenance of functional and species metrics.
Based on expectation due to human impacts, we expect to find significant functional changes, often accompanied by species change.

@@

We obtained mammal and bird time series from the BioTIME database, a global repository of high quality assemblage time series.
All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
All time series include abundance of observed species.
Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.

@@

All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
For variables with multiple levels each level was included as a separate trait axis.
Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
Metrics incorporated weighting based on species abundance.

@@

```{r resultsTab, echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything(), -REML, -bobyqa) %>%
  select(-c(n_ts, formula, lower.CL, upper.CL, significant)) %>%
  filter(group != "Residual" | is.na(group)) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
         metric = ifelse(metric == "S", "log(Species Richness)", metric),
         effect = ifelse(effect == "ran_pars", "random", effect),
         group = case_when(
           group == "rarefyID:study_id" ~ "time series within study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "SD Intercept",
           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
           term == "sd__year_scaled" ~ "SD Year",
           term == "sd__Observation" ~ "SD Observation",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2),
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
  dplyr::rename(grouping = group) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.") %>%
  column_spec(3, width = "2cm") %>%
  collapse_rows(columns = 1:3, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.

@@

indv_mod_table %>%
  filter(term == "year_scaled", p.value < 0.05) %>%
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
  mutate(positive = ifelse(estimate > 0, "+", "-")) %>%
  group_by(metric, positive) %>%
  summarise(n_study_id = n_distinct(study_id)) %>%
  pivot_wider(names_from = "metric", values_from = "n_study_id", values_fill = 0) %>%
  select(positive, S, Jaccard_base, FRic, FEve, FDiv, SES_FRic, SES_FEve, SES_FDiv) %>%
  kable(format = "latex", 
        col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.", 
        booktabs = TRUE, align = "c", escape = FALSE) %>%
  kable_styling("striped")

@@

At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
Around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover, but did show a significant shift in at least one summary functional metric.
These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continued adding to turnover.
Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.

@@

Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
Species are not therefore being replaced by exact functional counter parts.
The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss, but can also lead to increased functional richness.
While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the unavailability of an analog species in the species pool.

```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}

@@

Still, some results stand in contrast to predictions for trait shifts under global change.
For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
Instead, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].

@@

We present here four prevailing scenarios of change experienced in bird and mammal communities.
While we offer an first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.
It will be particularly critical to link forms of change to individual drivers to assess which drivers may impact species and functional diversity differently.
Understanding those links will help identify where directly measuring functional structure instead of just species change is necessary for understand impacts on a given system.
In addition, direct links to drivers will improve our ability to distinguish between communities experiencing no change due to a lack of perturbation from communities with high resilience in the face of disturbance.

>>>>

  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics. Using generalized linear mixed models, we estimated general trends in those metrics and trends for individual studies.

  Results: We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover. The lack of trend held even after correcting for changes in species richness. At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however most studies showed a shift in a species or functional metric.

  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window. With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.

@@

The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not considering the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
Still, there is general consensus on a few local-scale patterns that seem to be characteristic of the anthropocene, including the prevalence of species turnover across communities and patterns of both richness increase and decrease in different taxa and biomes [@dornelas2023].
Understanding the implications of change in species assemblage for ecosystem stability and function is a critical next step in moving forward the biodiversity change conversation.

@@

Here we perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
We expect to find significant functional changes happening alongside and independent of species richness changes as a result of ongoing anthropogenic impacts and global change drivers.

@@

We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series.
All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
All time series include abundance of observed species.
Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.

@@

All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
For variables with multiple levels each level was included as a separate trait axis.
Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
Metrics incorporated weighting based on species abundance.

@@

```{r resultsTab, echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything(), -REML, -bobyqa) %>%
  select(-c(n_ts, formula, lower.CL, upper.CL, significant)) %>%
  filter(group != "Residual" | is.na(group)) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
         metric = ifelse(metric == "S", "log(Species Richness)", metric),
         effect = ifelse(effect == "ran_pars", "random", effect),
         group = case_when(
           group == "rarefyID:study_id" ~ "time series within study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "SD Intercept",
           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
           term == "sd__year_scaled" ~ "SD Year",
           term == "sd__Observation" ~ "SD Observation",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2),
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
  dplyr::rename(grouping = group) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.") %>%
  column_spec(3, width = "2cm") %>%
  collapse_rows(columns = 1:3, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
  save_kable(here::here("figures/tables/results_table.jpeg"))
```

We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.

@@

indv_mod_table %>%
  filter(term == "year_scaled", p.value < 0.05) %>%
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
  mutate(positive = ifelse(estimate > 0, "+", "-")) %>%
  group_by(metric, positive) %>%
  summarise(n_study_id = n_distinct(study_id)) %>%
  pivot_wider(names_from = "metric", values_from = "n_study_id", values_fill = 0) %>%
  select(positive, S, Jaccard_base, FRic, FEve, FDiv, SES_FRic, SES_FEve, SES_FDiv) %>%
  kable(format = "latex", 
        col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.", 
        booktabs = TRUE, align = "c", escape = FALSE) %>%
  kable_styling("striped")%>%
  save_kable(here::here("figures/tables/trend_table.jpeg"))

@@

At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
In contrast to our expectation based on anthropogenic impacts, around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover, but did show a significant shift in at least one summary functional metric.
These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species-based metrics, particularly as species replacement continued adding to turnover.
Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.

@@

Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
Species are not therefore being replaced by exact functional counter parts.
The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss but can also lead to increased functional richness.
While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the inavailability of an analog species in the species pool.

```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}

@@

Still, some results stand in contrast to predictions for trait shifts under global change.
For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomenon which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
Instead, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].

@@

We present here four prevailing scenarios of change experienced in bird and mammal communities.
While we offer a first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.
It will be particularly critical to link forms of change to individual drivers to assess which drivers may impact species and functional diversity differently.
Understanding those links will help identify where directly measuring functional structure instead of just species change is necessary for understand impacts on a given system.
In addition, direct links to drivers will improve our ability to distinguish between communities experiencing no change due to a lack of perturbation from communities with high resilience in the face of disturbance.
